---
title: "Becker Running Analysis"
author: "Brad West"
date: "3/8/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      cache = TRUE,
                      fig.align = 'center')
knitr::opts_knit$set(root.dir = "~/drive/msu/spring2019/stat510/run")
options(scipen = 2, digits = 4)
```

```{r}
library(ggplot2)
library(magrittr)
```


```{r}
# Data paths
YBT_PATH <- "./data/YBT_SCRS.xlsx"
MUSCLE_DATA_SHEET <- "Muscle data"
RUNNING_DATA_SHEET <- "Running data"
UNITS_SHEET <- "Units"
# MUSCLE_DATA_PATH <- "../data/muscle_data.xlsx"

# muscle data colnames
legs <- c("r_leg", "l_leg")
measurements <- c("hip_abd", "hip_ext", "ext_rot", "int_rot")
metrics <- c("max", "mean")
units <- c("n", "kgs", "lbs")
muscle_data_cnames <- c("subject")
for (l in legs) {
  for (m in measurements) {
    for (t in metrics) {
      for (u in units) {
        muscle_data_cnames <- c(muscle_data_cnames, paste(l, m, t, u, sep = "_"))
      }
    }
  }
}
muscle_data_col_types <- c("text", 
                           rep("numeric", length(muscle_data_cnames) - 1))
```

```{r}
ybt_muscle <- readxl::read_excel(YBT_PATH, 
                                 sheet = MUSCLE_DATA_SHEET, 
                                 col_names = muscle_data_cnames,
                                 col_types = muscle_data_col_types,
                                 skip = 3)
ybt_running <- readxl::read_excel(YBT_PATH,
                                  sheet = RUNNING_DATA_SHEET,
                                  col_names = TRUE,
                                  skip = 2)
```

# Y-balance Metrics

There are three y-balance scores per leg (anterior reach, posterior-lateral reach, and posterior-medial reach), and a composite y-balance score which is derived from these three measurements.

We can plot the component measurements against the composite measurement, after standardizing for leg length.

```{r}
# y balance scores are likely dependent on height. Normalize by leg length
ybt_running <- ybt_running %>% 
  dplyr::mutate(L.A.norm = L.A.meters / L_leg_length,
                L.PL.norm = L.PL.meters / L_leg_length,
                L.PM.norm = L.PM.meters / L_leg_length,
                R.A.norm = R.A.meters / R_leg_length,
                R.PL.norm = R.PL.meters / R_leg_length,
                R.PM.norm = R.PM.meters / R_leg_length)
```

```{r}
# plot component y-balance against composite after standardizing for leg length
plot_component_vs_composite_y <- function(data, x, y, x_lbl = "x", y_lbl = "y", title = NULL) {
  if (is.null(title)) {
    title = paste(x_lbl, "vs", y_lbl)
  }
  p <- ggplot(data, mapping = aes_string(x, y)) +
    geom_point() +
    theme_bw() +
    ylab(y_lbl) +
    xlab(x_lbl) +
    labs(title = title)
  return(p)
}

component_vs_composite_plots <- list(
  c("L.A.norm", "L.CS", "Normalized L.A.meters", "L.CS", "Left leg Anterior Reach vs Composite - Normalized"),
  c("L.PL.norm", "L.CS", "Normalized L.PL.meters", "L.CS", "Left leg Posterior-Lateral vs Composite - Normalized"),
  c("L.PM.norm", "L.CS", "Normalized L.PM.meters", "L.CS", "Left leg Posterior-Medial vs Composite - Normalized"),
  c("R.A.norm", "R.CS", "Normalized R.A.meters", "R.CS", "Right leg Anterior Reach vs Composite - Normalized"),
  c("R.PL.norm", "R.CS", "Normalized R.PL.meters", "R.CS", "Right leg Posterior-Lateral vs Composite - Normalized"),
  c("R.PM.norm", "R.CS", "Normalized R.PM.meters", "R.CS", "Right leg Posterior-Medial vs Composite - Normalized")
)

for (cf in component_vs_composite_plots) {
  p <- plot_component_vs_composite_y(ybt_running, cf[1], cf[2], cf[3], cf[4], cf[5])
  print(p)
}
```

```{r}
# Get correlation between the normalized scores and non-normailzes scores
ybt_running_sml <- ybt_running %>% 
  dplyr::select(L.A.norm, R.A.norm, L.PM.norm, R.PM.norm, L.PL.norm, R.PL.norm, L.CS, R.CS)
corrplot::corrplot(cor(ybt_running_sml), method = "number", type = "upper", tl.col = "black", tl.srt = 45)
```


There is a much stronger correlation between the posterior measurements and the composite y-balance score than there is between the anterior measurements and the composite y-balance score.

```{r}
# Standard deviations of the measurements
apply(ybt_running_sml, 2, sd)
# Means of the measurements
apply(ybt_running_sml, 2, mean)
```

If we look at the individual variability of the measurements, we see that the standard deviation for the anterior measurements is half of that for the posterior measurements.
